Genome-wide changes in expression during inflorescence architecture development

### 2015-10-13 ms.Rmd
Stefan's comments 2015-10-13

---
title: Genome-wide changes in expression during inflorescence architecture development
  in rice
csl: csl/the-plant-journal.csl
header-includes: \usepackage{graphicx} \usepackage[utf8]{inputenc} \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  BiocStyle::html_document:
    dev: CairoPDF
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
    toc: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
  rmarkdown::tufte_handout: default
  word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
bibliography: bib/papersLib.bibtex
---

@@

At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA. Using strict cutoffs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries. As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))` and `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`), and four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridisation (**Figure Sx**).

@@

To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do]. Expressed genes were ranked by variance, and standardised, VST-transformed read counts for the 25% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw]. To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figures S", s_figCount, "sf_mfuzzCentroids"))` `r I(pasteLabel(" and S", s_figCount, "sf_mfuzzPca"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`). There are several expression patterns descibed by the clusters: genes that increase or decrease steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between RM and PBM and remain constant from PBM to SM (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`), genes that are constant between RM and PBM and then change between PBM and SM (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`). There were also two weaker clusters containing fewer genes, which had complex expression patterns involving changes in expression particular to the PBM (clusters `r getClusterName('inc', 'dec')` and `r getClusterName('dec', 'inc')`).

Examination of the clustered genes suggested that transcription factors were overrepresented in most clusters (**annotated cluster tables in SI**). To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests. Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database [@PerezRodriguez:2010co], a hypergeometric test was carried out if there were at least two members of one family in a given cluster. After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`). Cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the PBM–SM samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TFs whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 and G2-like TFs. MADS TFs and TRAF genes were overreprsented among genes that have low expression in the SM and PBM but higher expression in the RM (cluster `r getClusterName('small', 'inc')`) and bHLH, SBP and C2C2-GATA TFs were overreprented among genes with the inverse pattern (cluster `r getClusterName('small', 'dec')`)

Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). The latter,

Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')`. The LOG enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log* mutants produce small panicles with small SAMs and branching defects, and have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes [@Kurakawa:2007go]. In contrast, the LABA1 protein present in *O. rufipogon*, which has an identical primary sequence to the protein from *O. sativa* ssp. *japonica* cv. Nipponbare, also activates cytokinin *in vitro* [**ref Gu 2015; Hua et all 2015**], and introduction of the *Laba1* allele from *O. rufipogon* into an *O. sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [**Hua et all 2015**], but reduces the number of grains per panicle and the number of tillers per plant [**ref Gu 2015**], suggesting a link between increased cytokinin signalling and reduced panicle branching. *laba1* is a domestication allele found at high frequencies in cultivated accessions, suggesting that it was affected by artificial selection either for awn phenotype or grain yield [**ref Gu 2015; Hua et all 2015**]. No reports of *LOGL1* function in rice were found. An uncharecterized gene annotated as a cytokinin dehydrogenase, *LOC_Os05g31040*, is present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`), which suggests a model of increasing cytokinin activation and decreasing inactivation from RM–SM, but it is currently not clear how such a mechanism might influence meristem identity. There are also genes that may be involved in cytokinin signalling in other clusters, including a type-A response regulator (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')`, containing genes more highly expressed in RM, suggesting that the role of cytokinins in inflorescence development may be complex. **Grain number 1a?**. **Effect of cytokinin on meristem size/activity?**

The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes suggests that the clusters capture biologically meaningful groups of genes. Two G1-Like (G1L/ALOG) genes, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively. Only three G1L genes are expressed in the meristem RNA sequencing dataset, but they share a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`). Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff]. *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff]. This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5* [**This is just speculation!** I checked and there is some homology between the RNAi construct and other G1L genes, but I'm not sure how much is required for off-target effects.]

The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in flower development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions. Two other AP2 genes, *INDETERMINATE SPIKELET*, which acts redundantly with *SNB* to promote transition from SM to FM [@Lee:2012hj], and *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], were found in clusters containing genes that are highly expressed in the SM.

| **What could any of these patterns mean biologically?**
| **I picked out some genes from the clusters (list directly below), are there any others in here I should discuss?**

@@

The strong expression of SPL genes (which possess the SBP-box) in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development. Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. *WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc].

These patterns are not intended to represent each member of each gene family. For example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM. An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM. *PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *PAP2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz, @Kobayashi:2010hn]. In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples [**ref Jeon 2000, Agrawal 2005**].

@@

Briefly, adaptor trimming was performed with `cutadapt` [@Martin:2011eu] and reads were mapped against the MSUv7 genome and annotation [@Kawahara:2013gh; @Ouyang:2007il] downloaded from Phytozome 10.3 [@Goodstein:2012ho] using `STAR` [@Dobin:2012fg] in 2-pass mode. The `'quantMode'` argument of `STAR` was used for read counting. rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB [@Ohyanagi:2006go] or to regions where reads simulated with `wgsim` (https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped. Differential expression analysis to calculate Log~2~-fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do]. To determine a strict cutoff for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf]. These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distrubution to those used for calculating the 'genic' expression values. For each library, the cutoff was placed at the 95^th^ percentile of the distribution of intergenic expression values. Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package [@Kumar:2007uw] and geneset enrichment analysis with the `gage` package [@Luo:2009kp] respectively. All plots were produced with the `ggplot2` package [@Wickham:2009tn].

@@

**`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`**. Neighbour-joining similarity tree of putatative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*. Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers. The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems. Where there is no coloured label, the gene for the protein in question was removed from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do]. For rice, L~2~FC was estimated from the RM and SM samples reported here. For *A. thaliana*, L~2~FC was estimated using the inflorescence meristem and floret meristem samples reported by @Mantegazza:2014di. The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes. Clades are labelled following the nomenclature presented by (**ref: Arora paper**).

>>>>

---
title: Genome-wide changes in expression during inflorescence architecture development
  in rice
csl: csl/the-plant-journal.csl
header-includes: \usepackage{graphicx} \usepackage[utf8]{inputenc} \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  BiocStyle::html_document:
    dev: CairoPDF
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
    toc: yes
  word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
  pdf_document:
    fig_caption: yes
    number_sections: yes
  rmarkdown::tufte_handout: default
bibliography: bib/papersLib.bibtex
---

@@

At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA. Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries. As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))` and `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`), and four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridisation (**Figure Sx**). [**How do in situ / published results compare to transcriptome?**]

@@

To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do]. Expressed genes were ranked by variance, and standardised, VST-transformed read counts for the 25% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw]. To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figures S", s_figCount, "sf_mfuzzCentroids"))` `r I(pasteLabel(" and S", s_figCount, "sf_mfuzzPca"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`). Several common patterns were recovered by clustering: genes that increase or decrease in expression steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between RM and PBM and remain constant from PBM to SM (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`), genes that have constant expression between RM and PBM and then increase or decrease in expression between PBM and SM (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`). There were also two weaker clusters containing fewer genes, which had complex expression patterns involving changes in expression particular to the PBM (clusters `r getClusterName('inc', 'dec')` and `r getClusterName('dec', 'inc')`).

Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). Cluster `r getClusterName('dec', 'small')` contains three uncharacterized genes with annotations relating to auxin (*IAA10*, an auxin-responsive transcriptional regulator; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene *GH3-8*. [**Role of auxin influx / efflux in axillary / apical meristem specification**]. In *A. thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv]. Transgenic plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* (*i.e.* PBM and SBM; [@IkedaKawakatsu:2012co]), have smaller panicles with fewer branches [@Yadav:2011db]. GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that this phenotype may be correlated to reduced auxin signalling. [**Conclusion about auxin signalling**.]

**Auxin-related genes in clusters 2 & 4:**

+ *GH3-8*, *IAA10*, *ABCB27*, *LOC_Os01g63770* (AUX/LAX transporter) in cluster 2
+ *ARF6A*, *SAUR33*, *MDR12* (ABC), *ALS1* (ABC) in cluster 4

Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')` with genes that are increasingly expressed between RM and SM. The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants produce small panicles with small SAMs and branching defects, and have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes [@Kurakawa:2007go]. Reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield rice varieties [@Ashikari:2005eg]. These results suggest that an increase in active cytokinin is associated with grain yield via an effect on panicle branching, but the function of *LABA1* suggests the opposite. The LABA1 protein present in *O. rufipogon*, which has an identical primary sequence to the protein from *O. sativa* ssp. *japonica* cv. Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O. rufipogon* into an *O. sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy]. *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs]. No reports were found of *LOGL1* function, or the function of the gene annotated as a cytokinin dehydrogenase (*LOC_Os05g31040*) that is present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`), but the expression pattern of *LABA1* and these two genes suggests increasing cytokinin activation and decreasing inactivation from RM to SM. It is not currently clear how such a mechanism might influence meristem identity or panicle branching at the molecular level. [**Check for at/tomato papers**]. There are also genes that may be involved in cytokinin signalling in other clusters, including a type-A *RESPONSE REGULATOR* (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')`, containing genes more highly expressed in RM, suggesting that the role of cytokinins in inflorescence development may be complex.

Examination of the clustered genes suggested that transcription factors were overrepresented in most clusters (**annotated cluster tables in SI**). To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests. Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database [@PerezRodriguez:2010co], a hypergeometric test was carried out if there were at least two members of one family in a given cluster. After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`). Cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the PBM–SM samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TFs whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 and G2-like TFs. MADS TFs and TRAF genes were overrepresented among genes that have low expression in the SM and PBM but higher expression in the RM (cluster `r getClusterName('small', 'inc')`) and bHLH, SBP and C2C2-GATA TFs were overreprented among genes with the inverse pattern (cluster `r getClusterName('small', 'dec')`)

The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes. Only three G1-Like (G1L/ALOG) genes were detected in the dataset. Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`). Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff]. *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff]. This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*. [**Downstream targets of G1L5 (MADS22/MADS47/MADS55)**]

The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in panicle development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions. Two other AP2 genes, *INDETERMINATE SPIKELET*, which acts redundantly with *SNB* to promote transition from SM to FM [@Lee:2012hj], and *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], were found in clusters containing genes that are highly expressed in the SM.

| **I picked out some genes from the clusters (list directly below), are there any others in here I should discuss?**

@@

The strong expression of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development. Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. **Downstream targets of WFP**. *WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc].

These patterns are not intended to represent each member of each gene family. For example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM. An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM. *PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn]. In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples.

@@

Briefly, adaptor trimming was performed with `cutadapt` [@Martin:2011eu] and reads were mapped against the MSUv7 genome and annotation [@Kawahara:2013gh; @Ouyang:2007il] downloaded from Phytozome 10.3 [@Goodstein:2012ho] using `STAR` [@Dobin:2012fg] in 2-pass mode. The `'quantMode'` argument of `STAR` was used for read counting. rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB [@Ohyanagi:2006go] or to regions where reads simulated with `wgsim` (https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped. Differential expression analysis to calculate Log~2~-fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do]. To determine a strict cut-off for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf]. These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distribution to those used for calculating the 'genic' expression values. For each library, the cut-off was placed at the 95^th^ percentile of the distribution of intergenic expression values. Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package [@Kumar:2007uw] and geneset enrichment analysis with the `gage` package [@Luo:2009kp] respectively. All plots were produced with the `ggplot2` package [@Wickham:2009tn].

@@

**`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`**. Neighbour-joining similarity tree of putatative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*. Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers. The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems. Where there is no coloured label, the gene for the protein in question was removed from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do]. For rice, L~2~FC was estimated from the RM and SM samples reported here. For *A. thaliana*, L~2~FC was estimated using the inflorescence meristem and floret meristem samples reported by @Mantegazza:2014di. The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes. Clades are labelled following the nomenclature presented by [@Arora:2007fa].
